Arrows for Parallel Computation

### 2019-12-28 paper/content/response.md

XXXXXXXXXXXXXXXXX

> Another potential piece of related work is Turon's Reagents, that are
> based on an Arrow-like interface (albeit for concurrency, not
> parallelism): https://people.mpi-sws.org/~turon/reagents.pdf
>

Thank you for this reference, we include it now.

> ### Page 5
> 
> Para 1: Line 2: "which is basically"; better "A Haskell-like language
> for the JVM"

We adapted your wording.

> 
> Para 2; line 2: "typical GUI tasks" => you could be less vague.

> 
> Sec 3.1, para 1, last line: why not use the typeclass constraint
> "Monad m =>" instead of saying "where m is a Monad"?

> Sec 3.1, para 2, first line: What does the "This" at the end of the
> line refer to? Also, there is no mention of the Arrow laws.
>

> Sec 3.1, para 3: "syntactic sugar". I don't think these derived
> combinators should be called syntactic sugar, because you aren't
> defining the syntax of a language anywhere. They are either "derived
> combinators", or more prosaically, "utility functions".

We changed the way we reference those combinators throughout the text.
> 
> ### Page 6
> 
> The diagramatic representations of arrows on this page and the next
> are nice, but they don't seem to be used anywhere else. The diagrams
> in the text seem to be a mixture of communication topologies, mappings
> from input to output locations, and these arrow diagrams.
> 
> `ArrowApply`, `ArrowLoop` and `ArrowChoice` aren't defined anywhere -- how much
> background knowledge about Arrows do you expect the reader to have?

XXXXXXXXXXXXXXXXXXX

> "we have a truly more general interface as compared to a monadic one"
> -- but this isn't true for Eden right? And there are no examples in
> the paper of using non `(->)` arrows to write programs, unless I've
> missed one?

TODO: monadic RabinMiller!

> 
> ### Page 7
> 
> You need some reference to where Profunctors are defined, or at least
> a short introduction to them.

Fixed, as mentioned above.

> 
> "Among the most important are probably" ... "probably" is a vague. You
> could just say "the ones we focus on are...". Also, this sentence is
> sort of claiming that you chose these ones because they are the "most
> important", but the first sentence on the next page gives reasons in
> terms of the features of each one.
>

We have severely shortened and rewritten this paragraph alltogether, now there is no mention of importance at this point.

> ### Page 9
> 
> You could write `parEvalN` for the Par monad more concisely as:
> 
>    parEvalN fs as = runPar $ mapM (return >>> spawn >>= get) $ zipWith ($) fs as

Thank you very much, we replace the definition.

> 
> ### Page 10
> 
> You don't explain what the `Trans` typeclass is. `Trans`portable?

Yes, specialization of packing and unpacking and network transmission, e.g. streams for lists. We added two sentences explaining this issue.

> 
> ### Page 11
> 
> Why not put the `evalN` definition in the text for the reader?
> 
> ### Page 12
> 
> Figure 7: why not use a `newtype` for `Conf`?
> 
> Why is `Conf` for GpH in the figure (Fig 7), but not for the Par Monad
> instance (Fig 8)?
> 
> ### Page 13
> 
> Figure 9: does the second instance need the `Trans b` constraint?
> Should the `conf` before the `where` be a `Conf`?
>

XXXXXXXXXXXXXXXXXXX

> ### Page 14
> 
> Figure 11: what is the `ChunkSize` type? `Int`?

Yes. We added an explanation.

> 
> Where is the `ArrowApply` constraint used? Is it in `chunksOf` -- I
> can't find a definition of this function in the paper.
> 
> Section 4.3.2: the readbility of this subsection would be greatly
> improved by having the code of the function you are talking about!
> 
> ### Page 16
> 
> As I said above, Section 5 doesn't really explain how Futures fix the
> problem.

XXXXXXXXXXXXXX

> 
> ### Page 17
> 
> Fig 17, caption: "Other than" => "Unlike"

> 
> ### Page 18
> 
> second last line: remove the final "the"

> 
> ### Page 19
> 
> first line: remove "when"

> 
> ### Page 20
> 
> paragraph 3: "probably": do some benchmarks?

This was more of an speculation on what is the better in the API than a solid performance claim. We adjusted the wording correspondingly.

> 
> last line: "send" => "sent"

> 
> ### Page 22
> 
> "This combinator can, for example, be used to calculate the shortest
> paths in a graph using Warshall's algorithm." Have you done this?
> Could it be included in the benchmarks?
> 
> ### Page 23
> 
> Figure 27 is where the code starts to get really unreadable. If you
> are going to stick with Arrows, then why not use Paterson's arrow
> notation? Or give a diagramatic representation to help the reader?
> 
> ### Page 25
> 
> "We disregarded the hyper-threading ability in most of the cases."
> There should be a forward reference to the section discussing hyper
> threading specifically.
>

XXXXXXXXXXXXXXXXXX

> ### Page 37
> 
> last line: "as" => "was"

> 
> ### Page 38
> 
> The type signature for `(|||)` seems messed up (the ArrowChoice
> typeclass constraint is muddled up with the type of the first
> argument).

TODO TODO TODO

### 2019-12-28 paper/content/arxiv/ms.tex
reviewer's comments (3)

\paragraph{Arrows in other languages.}
Although this work is centred on Haskell implementation of Arrows, it is applicable to any functional programming language where parallel evaluation and Arrows can be defined. Preliminary experiments show that basic definitions of PArrows are possible in the Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (which is basically Haskell on the JVM). However, they are beyond the scope of this work, as are similar experiments with the Eta language\footnote{Eta project page at \url{http://eta-lang.org}}, 
a new approach to Haskell on the JVM.

\citet{achten2004arrows,achten2007arrow} use an Arrow implementation in Clean for better handling of typical GUI tasks. \citet{Dagand:2009:ORD:1481861.1481870} used Arrows in OCaml in the implementation of a distributed system.

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
        \section{Background}
	\label{sec:background}
	This section gives a short overview of Arrows
        (Section~\ref{sec:arrows}) and of GpH, the \ensuremath{\Conid{Par}} Monad, and
        Eden, the three parallel Haskells which we base our DSL on
        (Section~\ref{sec:parallelHaskells}).
	\subsection{Arrows}
\label{sec:arrows}
Arrows were introduced by \citet{HughesArrows} as a general interface for computation and a less restrictive generalisation of Monads. \citeauthor{HughesArrows} motivates the broader interface of Arrows with the example of a parser with added static meta-information that can not satisfy the monadic bind operator \ensuremath{(\bind )\mathbin{::}\Varid{m}\;\Varid{a}\to (\Varid{a}\to \Varid{m}\;\Varid{b})\to \Varid{m}\;\Varid{b}} (with \ensuremath{\Varid{m}} being a Monad)\footnote{In the example a parser of the type \ensuremath{\Conid{Parser}\;\Varid{s}\;\Varid{a}} with static meta information \ensuremath{\Varid{s}} and result \ensuremath{\Varid{a}} is shown to not be able to use the static information \ensuremath{\Varid{s}} without applying the monadic function \ensuremath{\Varid{a}\to \Varid{m}\;\Varid{b}}. With Arrows this is possible.}.

@@

\begin{figure}[h]
	\centering
	\begin{tabular}{cc}
		% \subcaptionbox
{\label{t1}}{\includegraphics[width = 1.5in]{first}} &
		% \subcaptionbox
{\label{fig:secondImg}}{\includegraphics[width = 1.5in]{second}} \\
\ensuremath{\Varid{first}} & \ensuremath{\Varid{second}} \\
\midrule
		% \subcaptionbox
{}{\includegraphics[width = 1.5in]{starstarstar}} &
		% \subcaptionbox
{}{\includegraphics[width = 1.5in]{andandand}}\\
\ensuremath{(\mathbin{*\!*\!*})}\label{fig:***Img} & \ensuremath{(\mathbin{\&\!\&\!\&})} \label{fig:&&&Img} \\
	\end{tabular}
	\caption{Visual depiction of syntactic sugar for Arrows.}
	\label{fig:syntacticSugarArrows}
\end{figure}
Hughes also defined some syntactic sugar (Fig.~\ref{fig:syntacticSugarArrows}): \ensuremath{\Varid{second}}, \ensuremath{\mathbin{*\!*\!*}} and \ensuremath{\mathbin{\&\!\&\!\&}}. 
\ensuremath{\Varid{second}} is the mirrored version of \ensuremath{\Varid{first}} (Appendix~\ref{utilfns}).
The \ensuremath{\mathbin{*\!*\!*}} function combines \ensuremath{\Varid{first}} and \ensuremath{\Varid{second}} to handle two inputs in one arrow, and is defined as follows:
% \begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}(\mathbin{*\!*\!*})\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{c}\;\Varid{d}\to \Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d}){}\<[E]%
\\
\>[B]{}\Varid{f}\mathbin{*\!*\!*}\Varid{g}\mathrel{=}\Varid{first}\;\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{second}\;\Varid{g}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% \caption{The (***) combinator}
% \label{fig:***}
% \end{figure}
The \ensuremath{\mathbin{\&\!\&\!\&}} combinator, which constructs an Arrow that outputs two different values like \ensuremath{\mathbin{*\!*\!*}}, but takes only one input, is:
% \begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}(\mathbin{\&\!\&\!\&})\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{a}\;\Varid{c}\to \Varid{arr}\;\Varid{a}\;(\Varid{b},\Varid{c}){}\<[E]%
\\
\>[B]{}\Varid{f}\mathbin{\&\!\&\!\&}\Varid{g}\mathrel{=}\Varid{arr}\;(\lambda \Varid{a}\to (\Varid{a},\Varid{a}))\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{*\!*\!*}\Varid{g}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% \caption{The (\&\&\&) combinator}
% \label{fig:&&&}
% \end{figure}
A~first short example given by Hughes on how to use Arrows is addition with Arrows:
%|add| over Arrows, which can be seen in Fig.~\ref{fig:addArrows}.
% \begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{add}\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Conid{Int}\to \Varid{arr}\;\Varid{a}\;\Conid{Int}\to \Varid{arr}\;\Varid{a}\;\Conid{Int}{}\<[E]%
\\
\>[B]{}\Varid{add}\;\Varid{f}\;\Varid{g}\mathrel{=}\Varid{f}\mathbin{\&\!\&\!\&}\Varid{g}\mathbin{>\!\!>\!\!>}\Varid{arr}\;(\lambda (\Varid{u},\Varid{v})\to \Varid{u}\mathbin{+}\Varid{v}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{NFData}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{parEvalN}\;\Varid{fs}\;\Varid{as}\mathrel{=}\Varid{runPar}\mathbin{\$}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}(\Varid{sequenceA}\;(\Varid{map}\;(\Varid{return}\mathbin{\circ}\Varid{spawn})\;(\Varid{zipWith}\;(\mathbin{\$})\;\Varid{fs}\;\Varid{as})))\bind \Varid{mapM}\;\Varid{get}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{parEvalN}\mathrel{=}\Varid{spawnF}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% \begin{figure}[h]
% 	\includegraphics[scale=0.5]{parEvalNEden}
% 	\caption{Dataflow of the Eden |parEvalN| version.}
% 	\label{fig:parEvalNEden}
% \end{figure}
% A simplistic graphical depiction of this definition can be found in Fig.~\ref{fig:parEvalNEden}.

@@

\subsubsection{Lazy \ensuremath{\Varid{parEvalN}}}
\begin{figure}[tb]
	\includegraphics[scale=0.7]{parEvalNLazy}
	\caption{\ensuremath{\Varid{parEvalNLazy}} depiction.}
	\label{fig:parEvalNLazyImg}
\end{figure}
The function \ensuremath{\Varid{parEvalN}} fully traverses the list of passed Arrows as well as their inputs. Sometimes this might not be feasible, as it will not work on infinite lists of functions like \eg \ensuremath{\Varid{map}\;(\Varid{arr}\mathbin{\circ}(\mathbin{+}))\;[\mskip1.5mu \mathrm{1}\mathinner{\ldotp\ldotp}\mskip1.5mu]} or just because we need the Arrows evaluated in chunks. \ensuremath{\Varid{parEvalNLazy}} (Figs.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this. It works by first chunking the input from \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} to \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{a}\mskip1.5mu]\mskip1.5mu]} with the given \ensuremath{\Varid{chunkSize}} in \ensuremath{\Varid{arr}\;(\Varid{chunksOf}\;\Varid{chunkSize})}. These chunks are then fed into a list \ensuremath{[\mskip1.5mu \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]\mskip1.5mu]} of chunk-wise parallel Arrows with the help of our lazy and sequential \ensuremath{\Varid{evalN}}. The resulting \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{b}\mskip1.5mu]\mskip1.5mu]} is lastly converted into \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{arr}\;\Varid{concat}}.
\begin{figure}[t]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{parEvalNLazy}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{conf}\to \Conid{ChunkSize}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to (\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]){}\<[E]%
\\
\>[B]{}\Varid{parEvalNLazy}\;\Varid{conf}\;\Varid{chunkSize}\;\Varid{fs}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{arr}\;(\Varid{chunksOf}\;\Varid{chunkSize})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{evalN}\;\Varid{fchunks}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{arr}\;\Varid{concat}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\mathbf{where}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\Varid{fchunks}\mathrel{=}\Varid{map}\;(\Varid{parEvalN}\;\Varid{conf})\;(\Varid{chunksOf}\;\Varid{chunkSize}\;\Varid{fs}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$ %% formatting
\caption{Definition of \ensuremath{\Varid{parEvalNLazy}}.}
\label{fig:parEvalNLazy}
\end{figure}

@@

Extensive use of \ensuremath{\Varid{pipe2}} over \ensuremath{\Varid{pipe}} with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to \ensuremath{\Varid{parEvalN}} inside of \ensuremath{\Varid{evalN}}. Nonetheless, we can define a parallel piping operator \ensuremath{\mathbin{|\!>\!\!>\!\!>\!|}}, which is semantically equivalent to \ensuremath{\mathbin{>\!\!>\!\!>}} similarly to other parallel syntactic sugar from Appendix~\ref{syntacticSugar}.

@@

\subsubsection{Ring skeleton} \label{sec:ring}
\begin{figure}[tb]
	\includegraphics[scale=0.75]{ring}
	\caption{\ensuremath{\Varid{ring}} skeleton depiction.}
	\label{fig:ringImg}
\end{figure}
Eden comes with a ring skeleton\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}} (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function \ensuremath{[\mskip1.5mu \Varid{i}\mskip1.5mu]\to [\mskip1.5mu \Varid{o}\mskip1.5mu]} with a ring of nodes that communicate with each other. Its input is a node function \ensuremath{\Varid{i}\to \Varid{r}\to (\Varid{o},\Varid{r})} in which \ensuremath{\Varid{r}} serves as the intermediary output that gets send to the neighbour of each node. This data is sent over direct communication channels, the so called \enquote{remote data}. We depict it in Appendix, Fig.~\ref{fig:ringEden}.
%\end{code}
%with toRD (to make use of remote data)
%\begin{code}
%\end{code}
%and rightRotate:
%\begin{code}

>>>>

\paragraph{Arrows in other languages.}
Although this work is centred on Haskell implementation of Arrows, it is applicable to any functional programming language where parallel evaluation and Arrows can be defined. Preliminary experiments show that basic definitions of PArrows are possible in the Frege language\footnote{GitHub project page at \url{https://github.com/Frege/frege}} (a Haskell-like language for the JVM). However, they are beyond the scope of this work, as are similar experiments with the Eta language\footnote{Eta project page at \url{http://eta-lang.org}}, 
a new approach to Haskell on the JVM.

\citet{achten2004arrows,achten2007arrow} use an Arrow implementation in Clean for graphical user interfaces. \citet{Dagand:2009:ORD:1481861.1481870} used Arrows in OCaml in the implementation of a distributed system.

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
        \section{Background}
	\label{sec:background}
	This section gives a short overview of Arrows
        (Section~\ref{sec:arrows}) and of GpH, the \ensuremath{\Conid{Par}} Monad, and
        Eden, the three parallel Haskells which we base our DSL on
        (Section~\ref{sec:parallelHaskells}).
	\subsection{Arrows}
\label{sec:arrows}
Arrows were introduced by \citet{HughesArrows} as a general interface for computation and a less restrictive generalisation of Monads. \citeauthor{HughesArrows} motivates the broader interface of Arrows with the example of a parser with added static meta-information that can not satisfy the monadic bind operator \ensuremath{(\bind )\mathbin{::}\Conid{Monad}\;\Varid{m}\Rightarrow \Varid{m}\;\Varid{a}\to (\Varid{a}\to \Varid{m}\;\Varid{b})\to \Varid{m}\;\Varid{b}}.\footnote{In the example a parser of the type \ensuremath{\Conid{Parser}\;\Varid{s}\;\Varid{a}} with static meta information \ensuremath{\Varid{s}} and result \ensuremath{\Varid{a}} is shown to not be able to use the static information \ensuremath{\Varid{s}} without applying the monadic function \ensuremath{\Varid{a}\to \Varid{m}\;\Varid{b}}. With Arrows this is possible.}.

@@

\begin{figure}[h]
	\centering
	\begin{tabular}{cc}
		% \subcaptionbox
{\label{t1}}{\includegraphics[width = 1.5in]{first}} &
		% \subcaptionbox
{\label{fig:secondImg}}{\includegraphics[width = 1.5in]{second}} \\
\ensuremath{\Varid{first}} & \ensuremath{\Varid{second}} \\
\midrule
		% \subcaptionbox
{}{\includegraphics[width = 1.5in]{starstarstar}} &
		% \subcaptionbox
{}{\includegraphics[width = 1.5in]{andandand}}\\
\ensuremath{(\mathbin{*\!*\!*})}\label{fig:***Img} & \ensuremath{(\mathbin{\&\!\&\!\&})} \label{fig:&&&Img} \\
	\end{tabular}
	\caption{Visual depiction of derived combinators for Arrows.}
	\label{fig:syntacticSugarArrows}
\end{figure}
Hughes also derived some further combinators (Fig.~\ref{fig:syntacticSugarArrows}): \ensuremath{\Varid{second}}, \ensuremath{\mathbin{*\!*\!*}} and \ensuremath{\mathbin{\&\!\&\!\&}}. 
\ensuremath{\Varid{second}} is the mirrored version of \ensuremath{\Varid{first}} (Appendix~\ref{utilfns}).
The \ensuremath{\mathbin{*\!*\!*}} function combines \ensuremath{\Varid{first}} and \ensuremath{\Varid{second}} to handle two inputs in one arrow, and is defined as follows:
% \begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}(\mathbin{*\!*\!*})\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{c}\;\Varid{d}\to \Varid{arr}\;(\Varid{a},\Varid{c})\;(\Varid{b},\Varid{d}){}\<[E]%
\\
\>[B]{}\Varid{f}\mathbin{*\!*\!*}\Varid{g}\mathrel{=}\Varid{first}\;\Varid{f}\mathbin{>\!\!>\!\!>}\Varid{second}\;\Varid{g}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% \caption{The (***) combinator}
% \label{fig:***}
% \end{figure}
The \ensuremath{\mathbin{\&\!\&\!\&}} combinator, which constructs an Arrow that outputs two different values like \ensuremath{\mathbin{*\!*\!*}}, but takes only one input, is:
% \begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}(\mathbin{\&\!\&\!\&})\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Varid{b}\to \Varid{arr}\;\Varid{a}\;\Varid{c}\to \Varid{arr}\;\Varid{a}\;(\Varid{b},\Varid{c}){}\<[E]%
\\
\>[B]{}\Varid{f}\mathbin{\&\!\&\!\&}\Varid{g}\mathrel{=}\Varid{arr}\;(\lambda \Varid{a}\to (\Varid{a},\Varid{a}))\mathbin{>\!\!>\!\!>}\Varid{f}\mathbin{*\!*\!*}\Varid{g}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% \caption{The (\&\&\&) combinator}
% \label{fig:&&&}
% \end{figure}
A~first short example given by Hughes on how to use Arrows is addition with Arrows:
%|add| over Arrows, which can be seen in Fig.~\ref{fig:addArrows}.
% \begin{figure}[h]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{add}\mathbin{::}\Conid{Arrow}\;\Varid{arr}\Rightarrow \Varid{arr}\;\Varid{a}\;\Conid{Int}\to \Varid{arr}\;\Varid{a}\;\Conid{Int}\to \Varid{arr}\;\Varid{a}\;\Conid{Int}{}\<[E]%
\\
\>[B]{}\Varid{add}\;\Varid{f}\;\Varid{g}\mathrel{=}\Varid{f}\mathbin{\&\!\&\!\&}\Varid{g}\mathbin{>\!\!>\!\!>}\Varid{arr}\;(\lambda (\Varid{u},\Varid{v})\to \Varid{u}\mathbin{+}\Varid{v}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{parEvalN}\;\Varid{fs}\;\Varid{as}\mathrel{=}\Varid{runPar}\mathbin{\$}\Varid{mapM}\;(\Varid{return}\mathbin{>>>}\Varid{spawn}\bind \Varid{get})\mathbin{\$}\Varid{zipWith}\;(\mathbin{\$})\;\Varid{fs}\;\Varid{as}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{parEvalN}\mathbin{::}(\Conid{Trans}\;\Varid{a},\Conid{Trans}\;\Varid{b})\Rightarrow [\mskip1.5mu \Varid{a}\to \Varid{b}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{b}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{parEvalN}\mathrel{=}\Varid{spawnF}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
% \begin{figure}[h]
% 	\includegraphics[scale=0.5]{parEvalNEden}
% 	\caption{Dataflow of the Eden |parEvalN| version.}
% 	\label{fig:parEvalNEden}
% \end{figure}
% A simplistic graphical depiction of this definition can be found in Fig.~\ref{fig:parEvalNEden}.
Here, the \ensuremath{\Conid{Trans}} typeclass originates from the Eden standard library and details the implementation of packing and of the transmission. A~default instance of \ensuremath{\Conid{Trans}} exists.

@@

\subsubsection{Lazy \ensuremath{\Varid{parEvalN}}}
\begin{figure}[tb]
	\includegraphics[scale=0.7]{parEvalNLazy}
	\caption{\ensuremath{\Varid{parEvalNLazy}} depiction.}
	\label{fig:parEvalNLazyImg}
\end{figure}
The function \ensuremath{\Varid{parEvalN}} fully traverses the list of passed Arrows as well as their inputs. Sometimes this might not be feasible, as it will not work on infinite lists of functions like \eg \ensuremath{\Varid{map}\;(\Varid{arr}\mathbin{\circ}(\mathbin{+}))\;[\mskip1.5mu \mathrm{1}\mathinner{\ldotp\ldotp}\mskip1.5mu]} or just because we need the Arrows evaluated in chunks. \ensuremath{\Varid{parEvalNLazy}} (Figs.~\ref{fig:parEvalNLazyImg},~\ref{fig:parEvalNLazy}) fixes this. It works by first chunking the input from \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]} to \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{a}\mskip1.5mu]\mskip1.5mu]} with the given \ensuremath{\Varid{chunkSize}} in \ensuremath{\Varid{arr}\;(\Varid{chunksOf}\;\Varid{chunkSize})}. These chunks are then fed into a list \ensuremath{[\mskip1.5mu \Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]\mskip1.5mu]} of chunk-wise parallel Arrows with the help of our lazy and sequential \ensuremath{\Varid{evalN}}. The resulting \ensuremath{[\mskip1.5mu [\mskip1.5mu \Varid{b}\mskip1.5mu]\mskip1.5mu]} is lastly converted into \ensuremath{[\mskip1.5mu \Varid{b}\mskip1.5mu]} with \ensuremath{\Varid{arr}\;\Varid{concat}}.
\begin{figure}[t]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{parEvalNLazy}\mathbin{::}(\Conid{ArrowParallel}\;\Varid{arr}\;\Varid{a}\;\Varid{b}\;\Varid{conf},\Conid{ArrowChoice}\;\Varid{arr},\Conid{ArrowApply}\;\Varid{arr})\Rightarrow {}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{conf}\to \Conid{ChunkSize}\to [\mskip1.5mu \Varid{arr}\;\Varid{a}\;\Varid{b}\mskip1.5mu]\to (\Varid{arr}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\;[\mskip1.5mu \Varid{b}\mskip1.5mu]){}\<[E]%
\\
\>[B]{}\Varid{parEvalNLazy}\;\Varid{conf}\;\Varid{chunkSize}\;\Varid{fs}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{9}{}\<[9]%
\>[9]{}\Varid{arr}\;(\Varid{chunksOf}\;\Varid{chunkSize})\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{evalN}\;\Varid{fchunks}\mathbin{>\!\!>\!\!>}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{arr}\;\Varid{concat}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\mathbf{where}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\Varid{fchunks}\mathrel{=}\Varid{map}\;(\Varid{parEvalN}\;\Varid{conf})\;(\Varid{chunksOf}\;\Varid{chunkSize}\;\Varid{fs}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
 %$ %% formatting
\caption{Definition of \ensuremath{\Varid{parEvalNLazy}}. The type \ensuremath{\Conid{ChunkSize}} is an alias to \ensuremath{\Conid{Int}}.}
\label{fig:parEvalNLazy}
\end{figure}

@@

We speculate that extensive use of \ensuremath{\Varid{pipe2}} over \ensuremath{\Varid{pipe}} with a hand-written combination data type would result in worse performance because of more communication overhead from the many calls to \ensuremath{\Varid{parEvalN}} inside of \ensuremath{\Varid{evalN}}. Nonetheless, we can define a parallel piping operator \ensuremath{\mathbin{|\!>\!\!>\!\!>\!|}}, which is semantically equivalent to \ensuremath{\mathbin{>\!\!>\!\!>}} similarly to other parallel derived combinators from Appendix~\ref{syntacticSugar}.

@@

\subsubsection{Ring skeleton} \label{sec:ring}
\begin{figure}[tb]
	\includegraphics[scale=0.75]{ring}
	\caption{\ensuremath{\Varid{ring}} skeleton depiction.}
	\label{fig:ringImg}
\end{figure}
Eden comes with a ring skeleton\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}} (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function \ensuremath{[\mskip1.5mu \Varid{i}\mskip1.5mu]\to [\mskip1.5mu \Varid{o}\mskip1.5mu]} with a ring of nodes that communicate with each other. Its input is a node function \ensuremath{\Varid{i}\to \Varid{r}\to (\Varid{o},\Varid{r})} in which \ensuremath{\Varid{r}} serves as the intermediary output that gets sent to the neighbour of each node. This data is sent over direct communication channels, the so called \enquote{remote data}. We depict it in Appendix, Fig.~\ref{fig:ringEden}.
%\end{code}
%with toRD (to make use of remote data)
%\begin{code}
%\end{code}
%and rightRotate:
%\begin{code}
